---
title             : "EDLD 651 Final Project: Discriminatory Experiences, Chronic Strain, Social Connectedness, and Psychological Wellbeing Among Individuals With Marginalized Sexual Orientations"
shorttitle        : "EDLD 651 Final Project"

author: 
  - name          : "Maggie Head"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "1215 University of Oregon, Eugene, OR 97403-1215"
    email         : "mhead@uoregon.edu"
  
  - name          : "Sarah Spafford"
    affiliation   : "1"

  - name          : "Heather Terral"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "University of Oregon"

authornote: |
  This study utilized data from Project STRIDE: Stress, Identity and Mental Health, which was funded by the National Institutes of Health/National Institute of Mental Health (Grant#: 5R01MH066058-03).

abstract: |
  Individuals with marginalized sexual orientations experience higher rates of physical and psychiatric comorbidities compared to their heterosexual counterparts. These disparities are considered the result of minority stress, such that the stress attached to navigating pervasive prejudice and discrimination precipitates deleterious mental health outcomes. Less is known about factors that are related to positive mental health outcomes in individuals with marginalized sexual orientations. Using data from 360 men and women with marginalized sexual orientations(i.e., lesbian, gay, bisexual, queer, or other LGB orientation) who participated in a three year longitudinal study in New York City, we examined the links between discriminatory experiences, chronic strain, social connectedness to the gay community, and psychological wellbeing. Results from a multiple regression analysis revealed discriminatory experiences and chronic strain were significantly negatively associated with psychological wellbeing. Consistent with hypotheses, social connectedness was significantly positively associated with psychological wellbeing. These findings provide further evidence for the relationship between minority stress and mental health and highlight the importance of social connectedness in promoting psychological wellbeing among LGBQ individuals.
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "minority stress, sexual minorities, LGBQ, health, social connectedness"
wordcount         : "X"

bibliography      : ["r-references.bib"]

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
header-includes:
  - \raggedbottom
  - \setlength{\parskip}{0pt}
output            : papaja::apa6_pdf
always_allow_html: true

---

```{r libraries, warning = FALSE}
library(papaja)
r_refs("r-references.bib")
library(haven)
library(rio)
library(janitor)
library(tidyverse)
library(apaTables)
library(qwraps2)
library(lm.beta)
library(psych)
library(lavaan)
library(lavaanPlot)
library(magick)
library(knitr)
```

```{r setup}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      eval=TRUE
                     )
```

# Introduction
  Inherent to living with a marginalized identity is the excess stress that accompanies stigma-related experiences and discriminatory conditions [@frost2015]. An extensive body of literature demonstrates that chronic exposure to stress compromises physical and mental health [see @thotis2010], and ultimately elevates susceptibility to a myriad of physiological and psychiatric disorders [@salleh2008]. It is not surprising, then, that individuals who identify as gay, bisexual, lesbian, and queer (LGBQ) experience higher rates of psychopathology than their heterosexual counterparts, including substance use disorders [@green2012], eating disorders [@parker2020], deliberate self-injury [@king2008], suicidality, and suicide attempts [@haas2010]. The term “minority stress” has been used to describe the phenomenon of elevated mental health concerns resulting from the societal stigmatization of LGBQ sexual orientation status [@meyer1995]. The link between minority stress and poor health outcomes may be direct, such that discriminatory experiences lead to increased cortisol [@korous2017] and cardiovascular reactivity [@panza2019]. However, minority stress may also impact health indirectly through the cognitive burden, strain, and behavioral coping strategies that are required to navigate marginalization [@frost2015].
  
  Given that morbidity and mortality is intimately tied to social and interpersonal conditions, researchers have come to recognize the importance of relationships and support [@cohen2004 ; @pescosolido2011]. Social connectedness, which refers to the sense of subjective belonging that people feel in relation to individuals and groups of others, is considered a pivotal factor in individual and population-level health [@haslam2015]. Burgeoning evidence indicates that, among individuals with marginalized identities, connection with others who are marginalized for the same characteristic may mitigate detrimental stress responses [@austin2017]. Indeed, social connectedness is associated with positive health outcomes and has been found to buffer the negative effects of discrimination and perceived stress among many groups of marginalized individuals[@kim2017; @liao2016; @liu2020]. Yet, social connectedness is markedly overlooked in research examining the health of LGBQ individuals. Thus, the purpose of the current study was to examine the longitudinal relationships between discriminatory experiences, chronic strain, social connectedness, and psychological wellbeing among LGBQ individuals. 

```{r data import, include = FALSE}
stride <- read_sav("data/project_stride_data.sav") %>% 
  as_tibble() %>% 
  janitor::clean_names()
head(stride)

stridy <- stride %>%
  drop_na(perwellb_2, dis_d_total, chr_gen_2, connect) %>% 
  filter(sexori <= 6) %>% 
    select(sexori,
           age,
           ethnic,
           gender,
           chr_gen_2,
           perwellb_2,
           dis_d_total,
           connect) %>%
  mutate(perwellb_2 = as.numeric(perwellb_2),
         chr_gen_2 = as.numeric(chr_gen_2),
         sexori = as.factor(sexori),
         gender = as.factor(gender),
         ethnic = as.factor(ethnic),
        ethnic = fct_recode(ethnic,
    "White" = "0",
    "Black/African-American" = "1",
    "Latino/Hispanic" = "2"), 
         sexori = fct_recode(sexori,
    "Gay" = "1",
    "Lesbian" = "2",
    "Queer" = "3",
    "Bisexual" = "4",
    "Homosexual" = "5",
    "Other - LGB" = "6"))
```

```{r tidy data, include = FALSE}
stridy_wide <-stridy %>%
  group_by(sexori, ethnic) %>%
  characterize() %>%
  summarize(mean_connect = mean(connect)) %>%
  pivot_wider(
    names_from = ethnic,
    values_from = mean_connect
  )

stridy_long <- stridy_wide %>%
  pivot_longer(cols = c(2:4),
               names_to = "Ethnicity",
               values_to = "Connectedness")
               

```

```{r demographics, include = FALSE}
count(stridy)
count(stridy, ethnic)
count(stridy, gender)
desc_sexori <- count(stridy, sexori)

stridy$age <- as.numeric(stridy$age)
mean(stridy$age, na.rm = TRUE)
sd(stridy$age, na.rm = TRUE)
range(stridy$age, na.rm = TRUE)
```




```{r means, include=FALSE}
mean_age <- mean(stridy$age, na.rm = TRUE)

sd_age <- sd(stridy$age, na.rm = TRUE)

tabyl(stridy$sexori, sort = TRUE) %>% 
  kable()
```
# Methods
## Participants
Project STRIDE [@projectstride] participants included individuals who had been residing in New York City for a minimum of two years, self-identified as lesbian, gay, bisexual (LGB), or straight, and self-identified as White, Black, or Latino [@projectstride]. Participants were excluded from the present study if they identified as heterosexual or did not complete the main study measures (_n_ = 360). Participants were aged 18-59 years (_M_ = `r mean_age`, _SD_ = `r sd_age`) and were predominantly White (34%), followed by Black/African-American (33%), and Latino/Hispanic (32%). The distribution of sexual orientations in the study sample can be seen in Table 1. 
\newpage
  __Table 1.__

_Distribution of self-identified sexual orientations_
```{r sexori table, results='asis'}
kable(desc_sexori,col.names = c('Sexual Orientation', 'Count'))
```

## Measures
### Discriminatory experiences
The discriminatory experiences 8-item measure was adapted from @williams1997 to be inclusive of all minority groups (e.g. gender minorities). This scale assessed how often discriminatory experiences (e.g. being treated with less respect, being threatened or harassed) occurred throughout their lifetimes. Each question was rated on a 4-point scale (1 = _"often"_ through 4 = _"never"_) and coded so that higher scores represented more discriminatory experiences [@projectstridemethod]. For these analyses, the total number of types (0-8) of everyday discrimination experiences were used.

### Chronic strain
The chronic strain measure was adapted from a scale by @wheaton1999, which measures strain in 9 areas of life, including general problems, financial issues, work relationships, parenting, family, social life, residence, and health. Responses were coded such that higher scores indicated higher levels of chronic strain [@projectstridemethod].

### Social connectedness
Social connectedness was contextualized as connectedness to the gay community, as measured by an 8-item scale adapted from @mills2001 to be more relevant to the geographic area. Each response was rated from 1 (_"agree strongly"_) to 4 (_"disagree strongly"_) and coded so that higher scores indicated a greater level of connectedness to the gay community [@projectstridemethod].

### Psychological wellbeing
Psychological wellbeing was assessed using an 18-item measure adapted from @ryff1989 and @ryffkeyes1995. This measured psychological wellbeing on six dimensions including dimensions of self-acceptance, purpose in life, environmental mastery, positive relations with others, personal growth, and autonomy. All responses were coded such that higher scores indicated higher levels of wellbeing [@projectstridemethod].


## Procedure
For additional details on data collection procedures for Project STRIDE, please see @projectstridemethod.

## Data Analytic Strategy and Hypotheses
Prior to the main analysis, data were screened for missingness. Pearson bivariate correlations were conducted among discriminatory experiences, chronic strain, social connectedness, and psychological wellbeing. To examine the proposed model, a multivariate regression analysis was conducted. Discriminatory experiences, chronic strain, and social connectedness were entered as the predictor variables. Psychological wellbeing was entered as the outcome variable. We expected a negative association between discriminatory experiences and psychological wellbeing (Hypothesis 1). We also expected a negative association between chronic strain and psychological wellbeing (Hypothesis 2). In contrast, we expected a positive association between social connectedness and psychological wellbeing (Hypothesis 3).

We used `r cite_r("r-references.bib")` for all our analyses.

# Results
## Preliminary Analyses
  Missing data were minimal; thus, listwise deletion was employed. Means, standard deviations, minimum and maximum values of the main study measures for the total sample can be seen in Table 2. Means, standard deviations, minimum and maximum values of the main study variables according to sexual orientation can be seen in Table 3. Of particular concern is the substantial number of discriminatory experiences reported by participants. Figure 1 displays the average number of everyday discriminatory experiences according to sexual orientation. Pearson bivariate correlations revealed small to moderate correlations among the main study variables (see Figure 2).

 \newpage
 __Table 2.__

_Descriptive statistics for main study variables._
```{r summary table, results= "asis"}
options(qwraps2_markup = "markdown")
summary1 <-
  list("Everyday Discrmination" =
         list("min"       = ~ round(min(dis_d_total, na.rm = TRUE),2),
              "median"    = ~ round(median(dis_d_total, na.rm = TRUE),2),
              "max"       = ~ round(max(dis_d_total, na.rm = TRUE),2),
              "mean (sd)" = ~ qwraps2::mean_sd(dis_d_total, na_rm = TRUE)),
       "Chronic Strain" =
         list("min"       = ~ round(min(chr_gen_2, na.rm = TRUE),2),
              "median"    = ~ round(median(chr_gen_2, na.rm = TRUE),2),
              "max"       = ~ round(max(chr_gen_2, na.rm = TRUE),2),
              "mean (sd)" = ~ qwraps2::mean_sd(chr_gen_2, na_rm = TRUE)),
       "Psychological Wellbeing" =
         list("min"       = ~ round(min(perwellb_2, na.rm = TRUE),2),
              "median"    = ~ round(median(perwellb_2, na.rm = TRUE),2),
              "max"       = ~ round(max(perwellb_2, na.rm = TRUE),2),
              "mean (sd)" = ~ qwraps2::mean_sd(perwellb_2, na_rm = TRUE)),
       "Social Connectedness" =
         list("min"       = ~ round(min(connect, na.rm = TRUE),2),
              "median"    = ~ round(median(connect, na.rm = TRUE),2),
              "max"       = ~ round(max(connect, na.rm = TRUE),2),
              "mean (sd)" = ~ qwraps2::mean_sd(connect, na_rm = TRUE))
  )
table_total <- summary_table(stridy, summary1)
table_total
```
\newpage
__Table 3.__

_Descriptive statistics for main study variables by sexual orientation_
```{r summary table 2, results="asis"}
grouped_table <- summary_table(stridy%>%
                                 mutate_at('sexori', characterize),
                               summaries = summary1, by = c("sexori"))
grouped_table
```

__Figure 1.__

_Experiences of everyday discrimination according to sexual orientation._

```{r meanplot, results="asis", fig.height=3, fig.width=5}
mean_plot <- stridy %>% 
  group_by(sexori) %>% 
  summarize(mean_disc = mean(dis_d_total)) %>% 
  ggplot(aes(sexori, mean_disc)) +
  geom_col(fill = "plum4", alpha = .75) +
  labs(title = "Experiences of Everyday Discrimination",
       subtitle = "According to Sexual Orientation") +
  xlab("Sexual Orientation") +
  ylab("Average Discriminatory Experiences")+
  theme(text=element_text(size=12,  family="serif"))

mean_plot
```
```{r labels}
cor_labels <- list(chr_gen_2 = "Chronic Strain", perwellb_2 = "Psychological Wellbeing", dis_d_total = "Discrimination", connect = "Social Connectedness")
```
\newpage
__Figure 2.__

_Correlation Panels and Distributions For All Variables Included in the Model._

```{r correlation panels, echo=FALSE, fig.height=5, fig.width=7}
cor_labels <- list(chr_gen_2 = "Chronic Strain", perwellb_2 = "Psychological Wellbeing", dis_d_total = "Discrimination", connect = "Social Connectedness")

pairs_plot <- pairs.panels(stridy[,-c(1:4)], 
             method = "pearson", 
             hist.col = "darkmagenta",
             density = FALSE,
             ellipses = FALSE,
             lm = TRUE, 
             labels = cor_labels)
```

## Primary Analyses
```{r regression, include=FALSE}
mod <- lm(perwellb_2 ~  1 + dis_d_total + chr_gen_2 + connect, data = stridy)
summary(lm.beta(mod))

apa.aov.table(mod)
```
A multiple regression analysis was conducted to examine the effects of discriminatory experiences, chronic strain, social connectedness on psychological wellbeing among LGBQ individuals. When all variables were entered into the model, discriminatory experiences were negatively associated with psychological wellbeing, $\hat{\beta_{1}}=-0.05, SE(\hat{\beta_{1}})=-0.11, t(356)=-2.14, p=.03$. Likewise, consistent with hypothesis 2, chronic strain was significantly negatively associated with psychological wellbeing, $\hat{\beta_{2}}=-0.29, SE(\hat{\beta_{2}})=-0.20, t(356)=-3.91, p < .001$. Consistent with hypothesis 3, social connectedness was significantly positively associated with psychological wellbeing, $\hat{\beta_{3}}=0.24, SE(\hat{\beta_{3}})=0.15, t(356)=2.99, p < .001$. Taken together, all three predictors explained approximately 7.7% of the variance in psychological wellbeing, $F(3,356)=9.90, p<.001, R^{2}=.077$. Figure 3 displays the relationship between everyday discrimination and psychological wellbeing. Figure 4 displays the path model with corresponding beta coefficients. 

__Figure 3.__

_Path model for the effect of discrimination, chronic strain, and social connectedness on psychological wellbeing._

```{r regression plot, echo=FALSE}
ggplot(stridy, aes(x = dis_d_total, y = perwellb_2 )) +
   geom_point(color = "orange") +
   geom_smooth(method = "lm") +
  labs(x = "Discrimination", y = "Pychological Wellbeing") +
  theme(text=element_text(size=12,  family="serif"))
```
```{r lavaan, include=FALSE}
model <- 'perwellb_2 ~ dis_d_total + chr_gen_2 + connect'

fit <- sem(model, data = stridy)
summary(fit)

labels <- list(perwellb_2 = "Psychological Wellbeing", dis_d_total = "Discrimination", chr_gen_2 = "Chronic Strain", connect = "Social Connectedness")
```
\newpage
__Figure 4.__

_Path model for the effect of discrimination, chronic strain, and social connectedness on psychological wellbeing._

```{r lavaan plot, echo=FALSE}
lavaanPlot(model = fit, labels = labels, node_options = list(shape = "box", fontname = "Garamond"), edge_options = list(color = "black"), coefs = T, stars = c("regress"))
```

# Discussion

Findings that social connectedness was positively associated with psychological well-being are simultaneously intuitive and in synchrony with the literature. For members of the LGB community, both physical and virtual spaces in which members can create and maintain meaningful relationships with each other may be one method to increase resilience within these groups.
(Discuss discrimination and strain, intersectionality)
In considering interventions to improve health and well-being among LGB communities, it is important to not place the burden on these groups; a preventative approach in line with a social ecological perspective must also include mechanisms designed to reduce both implicit and explicit bias in the  general population. Approval of LGBT+ individuals has decreased among young people in the United States ages 18 - 34 @Miller2019, and the presidential administration of 2016 - 2020 took several actions during their tenure which further marginalized this population @Acosta2020. Although this study did not measure social connectedness outside of the LGB community, future research examining perceived acceptance by the general population may better inform health promotion interventions for this group.


## Strengths and Limitations
The availability and tidyness of the data from this robust sample gathered by @projectstride are a clear strength of the present work. However, it is unclear whether or not these results are applicable to LGB members outside of New York City itself, and furthermore within rural areas. Future work may benefit from examining whether LGB individuals who relocate to large cities share similiar rates of discrimination and may be systematically different from community members who have no relocated.Additionally, the intentional exclusion of trans people from this sample limits the richness of this dataset and interpretability. 
\newpage

# References

```{=tex}
\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
```
::: {#refs custom-style="Bibliography"}
:::

\endgroup
